
Metrics and metacommunities and sequence read data exploration

Description:
1) Atlantic CABIN Sequence read plots
2) GRDI metacommunity data exploration (just the beginning, data exploration not finished)

files used: "Filtered_genus_level_data_MaritimeRivers_noblank.csv"
            "GRDI dataset - ALL - filtered - GENUS-LEVEL - 2022-05-11.csv"

#1) Sequence Reads 
```{r}

library(dplyr)
library(tidyverse)

#load in data
sequence_data <- read.csv("Filtered_genus_level_data_MaritimeRivers_noblank.csv", header = TRUE)

#calc esv sizes
calc_esvsizes <- sequence_data %>% filter(Order=='Trichoptera'|Order=='Plecoptera_Insecta'|Order=='Ephemeroptera') %>%
    group_by(GRDI_MixedSpecimenNumber, TrimmedPrimer, Genus, Order) %>%
    summarise(ESVsize = sum(ESVsize))

#calc species richness
esv_richness <- calc_esvsizes %>% 
  group_by(GRDI_MixedSpecimenNumber, TrimmedPrimer) %>%
  mutate("species_richness" = n()) %>%
  group_by(GRDI_MixedSpecimenNumber, TrimmedPrimer, species_richness) %>%
  summarise(ESVsize_sum = sum(ESVsize))

#correlation between esv and species richness for each primer 
for (i in unique(esv_richness$TrimmedPrimer)) {
  x <- esv_richness[esv_richness$TrimmedPrimer == i,]$ESVsize_sum
  y <- esv_richness[esv_richness$TrimmedPrimer == i,]$species_richness
  m <- cor(x, y)
  print(i)
  print(m)
}
#results: low positive correlation across the board

```


```{r}

#plot richness vs. esv for each primer

for (i in unique(esv_richness$TrimmedPrimer)) {

  ESVsize_sum <- esv_richness[esv_richness$TrimmedPrimer == i,]$ESVsize_sum
  species_richness <- esv_richness[esv_richness$TrimmedPrimer == i,]$species_richness
  Trimmed_Primer <- esv_richness[esv_richness$TrimmedPrimer == i,]$TrimmedPrimer[1] 
  plot(ESVsize_sum ~ species_richness)
  title(Trimmed_Primer)

}

```
#2) GRDI data exploration for metacommunities and metrics - unfinished, this is what is done so far. JL Sept 2024
```{r}

grdi_data <- read.csv('GRDI dataset - ALL - filtered - GENUS-LEVEL - 2022-05-11.csv')

```


```{r}

#How many unique genuses were detected in each sample? (for all primers)
grdi_data %>%
  group_by(GRDI_MixedSpecimenNumber, Genus) %>%
  select(GRDI_MixedSpecimenNumber, Genus) %>%
  distinct() %>%
  group_by(GRDI_MixedSpecimenNumber) %>% 
  tally(name = 'unique_genera_detected')

```


```{r}

#How many genera were detected in each sample, for each primer?
grdi_data %>%
  group_by(GRDI_MixedSpecimenNumber, Genus, TrimmedPrimer) %>%
  select(GRDI_MixedSpecimenNumber, Genus, TrimmedPrimer) %>%
  distinct() %>%
  group_by(GRDI_MixedSpecimenNumber, TrimmedPrimer) %>% 
  tally(name = 'genera_detected')

#these are not unique genera - ie. for sample, primer there could be the same genera detected >1x 

```

```{r}




```

